---
title: 'Tame Your Data With R: ADVANCE Workshop 2'
author: "Elizabeth E. Esterly"
date: "December 4th, 2017, University of New Mexico"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r tidyItUp}
library(tidyverse)
```
```{r DatasetIntoPlace}
super <- read_csv('superheroes.csv') 
```

##Transform
We'll bring in a second dataset here, transform it, and then transform our original dataset by doing a join on superhero names. For brevity's sake, assume this one is already tidied up. 
```{r bringInSecond, message=FALSE}
#Import the second set and take a look at the first few rows
second <- read_csv("~/Desktop/characteristics.csv")
second %>% head()
```
Huh. It looks like the name column repeats itself. Let's get a wider view...
```{r bringInSecond2, message=FALSE}
second
```

###spread and gather
It definitely does. This data is a great candidate for transformation. Let's break up that status column into separate factors:
```{r spread}
#Import the second set and take a look at the first few rows
secondWide <- second %>% spread(status, measurement)
secondWide
```

The Tidyverse `gather` function will help you do the opposite--go from wide to long. To learn more about that function, you can type:
```{r gather}
help("gather")
```
You can do this with most functions in R to get some more details about them.

Note that the result of `spread` was sorted alphabetically by the first column.
We really care about who's the fastest, though:
```{r sortby}
#Sort by speed
#ascending order
secondWide <- secondWide %>% arrange(flyingSpeedmph)
secondWide
#descending order
secondWide <- secondWide %>% arrange(desc(flyingSpeedmph))
secondWide
```

This looks like what we want. Let's merge with our `super` dataset.

###joins  
Don't be scared!  

####inner join(x, y)
From the documentation:
```{r helpagain}
help("inner_join")
```
>Return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches >between x and y, all combination of the matches are returned. This is a mutating join.

We'll give our `super` dataset as x, because we want to keep all of the characters in `super` that also appear in `secondWide`. `secondWide` will be given second as y.  
*What is a mutating join?*: It's some database terminology. It just means you can add variables to the LHS.
```{r inner}
#Try an inner join
inner_join(super, secondWide)
#The Tidyverse way: use pipes for first function argument 'x'
#Doesn't make a difference here but a good habit to get into
super %>% inner_join(secondWide)
```
Notice we were automatically joined on name. We kept all characters that appeared in *both* datasets and merged all attributes given in both.


####semi_join(x, y)  
>return all rows from x where there are matching values in y, keeping just columns from x
```{r semi}
super %>% semi_join(secondWide)
```
Here, our names were filtered by just the ones that appeared in both datasets, but we didn't pick up the extra columns from the `secondWide` dataset.

####anti_join(x, y)
>return all rows from x where there are not matching values in y, keeping just columns from x.
```{r anti}
super %>% anti_join(secondWide)
```
In this case, we exclude all characters that appear in both x and y but still keep just columns from x.  

####full_join(x, y)  
>return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing.
```{r full}
super %>% full_join(secondWide)
```
Let's do an anti join one more time and save the result:
```{r innerAgain}
result <- super %>% anti_join(secondWide)
result
```
Our dataset has come a long way since the beginning! Let's close by exploring some visualizations.


## Visualize

The `ggplot2` library provides an extensive set of tools to help you create beautiful visualizations. It can be hard to get started though, so I suggest looking at great examples, like from this website:
http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html#Ordered%20Bar%20Chart
and working backwards from there.  

### Categorical bar chart
Let's compare alignment between the sexes.
```{r align}
result %>% ggplot(aes(sex)) +
  geom_bar(aes(fill=alignment)) 
```

### Lollipop charts
```{r lollipop}
result %>% ggplot(aes(x=name, y=age, label=age )) + 
  geom_point(stat='identity', fill="black", size=8) +
  geom_segment(aes(y = 0, 
                   x = name, 
                   yend = age, 
                   xend= name), 
               color = "black") +
  coord_flip()
```
Loki is throwing things off here. Let's take him out:  
```{r lollipop2}
result %>% filter(name != "Loki") %>% ggplot(aes(x=name, y=age, label=age )) + 
  geom_point(stat='identity', fill="black", size=8) +
  geom_segment(aes(y = 0, 
                   x = name, 
                   yend = age, 
                   xend= name), 
               color = "black") +
  coord_flip()
```
Much better!

Let's add a little text in the dots for clarity and set our 'baseline' age to 32:
```{r lollipop3}
result %>% filter(name != "Loki") %>% ggplot(aes(x=name, y=age, label=age )) + 
  geom_point(stat='identity', fill="black", size=8) +
  geom_segment(aes(y = 32, 
                   x = name, 
                   yend = age, 
                   xend= name), 
               color = "black") +
                geom_text(color="white", size=3)+
  labs(title="Who's older and younger than the Batman?", subtitle="Batman is 32")+
  coord_flip()
```