---
title: 'Tame Your Data With R: ADVANCE Workshop 2'
author: "Elizabeth E. Esterly"
date: "December 4th, 2017, University of New Mexico"
output:
  html_document: default
  pdf_document: default
---
##Getting Started
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Download these datasets to your desktop.
DATASET 1: http://cs.unm.edu/~elizabeth/newSuperheroes.csv  
DATASET 2: http://cs.unm.edu/~elizabeth/newCharacteristics.csv  

```{r tidyItUp, message=FALSE}
#Load the Tidyverse
library(tidyverse)
```
```{r DatasetIntoPlace, message=FALSE}
#Read in the superhero data
super <- read_csv('newSuperheroes.csv') 
```

##Transform
We'll bring in a second dataset here, transform it, and then transform our original dataset by doing some different joins of these two datasets.
```{r bringInSecond, message=FALSE}
#Import the second set and take a look at the first few rows
second <- read_csv('newCharacteristics.csv')
second %>% head()
```
Huh. It looks like the name column repeats itself. Let's get a more complete view...
```{r bringInSecond2, message=FALSE}
second
```
I think we should do a transformation here, but let's do it after the join, so we can transform all at once. 

###joins  
Don't be scared! Take it slow. 

####inner join(x, y)
From the documentation:
```{r helpagain}
help("inner_join")
```
>Return all rows from x where there are matching values in y, and all columns from x and y. If there are multiple matches >between x and y, all combination of the matches are returned. This is a mutating join.

We'll give our `super` dataset as x, because we want to keep all of the characters in `super` that also appear in `second`. `second` will be given second as y.  
*What is a mutating join?*: It's some database terminology. It just means you can add variables to the LHS.
```{r inner}
#Try an inner join
inner_join(super, second)
#The Tidyverse way: use pipes for first function argument 'x'
#Same result
superInner <- super %>% inner_join(second)
```
Notice we were automatically joined on name. We kept all characters that appeared in *both* datasets and merged all attributes given in both.

###spread and gather
This data is a great candidate for transformation. Let's break up that status column into separate factors by using spread to make it a wide format:
```{r spread}
#Use the spread command to spread the status column out. (Go to "wide.")
superInnerSpread <- superInner %>% spread(status, measurement, convert=TRUE)
superInnerSpread
```

(We'll just have time to practice the spread function today, but the Tidyverse `gather` function will help you do the opposite--go from wide to long. To learn more about that function, you can type:
```{r gather}
help("gather")
```
You can do this with most functions in R to get some more details about them.)

Let's learn more joins and practice using the spread function to transform them.
####semi_join(x, y)  
>return all rows from x where there are matching values in y, keeping just columns from x
```{r semi}
superSemi <- super %>% semi_join(second)
superSemi
```
Here, our names were filtered by just the ones that appeared in both datasets, but we didn't pick up the extra columns from the `second` dataset like we did in the inner join. There's nothing to spread on. Let's try a full join now.

####full_join(x, y)  
>return all rows and all columns from both x and y. Where there are not matching values, returns NA for the one missing.
```{r full}
superFull <- super %>% full_join(second)
superFull
```

...and spread it out.
```{r fullSpread}
superFullSpread <- superFull %>% spread(status, measurement, convert=TRUE, fill="FILL ME!")
superFullSpread
```

####anti_join(x, y)
>return all rows from x where there are not matching values in y, keeping just columns from x.
```{r anti}
superAnti <- super %>% anti_join(second)
superAnti
```

In this case, we *exclude* all characters that appear in both x and y but still keep just columns from x.  
Let's close by exploring some visualizations.

## Visualize

The `ggplot2` library provides an extensive set of tools to help you create beautiful visualizations. It can be hard to get started though, so I suggest looking at great examples, like from this website:
http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html#Ordered%20Bar%20Chart
and working backwards from there.  

## QUESTION 1: WHO'S BAD?
### Categorical bar chart
Let's compare alignment between the sexes.
Who's more evil in our anti-joined result?
 
```{r alignAnti}
superAnti %>% ggplot(aes(sex)) +
  geom_bar(aes(fill=alignment)) 
```

How about our spread full-joined result? Our semi-join result? Our spread inner-joined result?
How can we handle NA's?

##QUESTION 2: WHO'S OLDER AND YOUNGER THAN BATMAN?
### Lollipop charts

```{r lollipop}
superAnti %>% ggplot(aes(x=name, y=age, label=age )) + 
  geom_point(stat='identity', fill="black", size=8) +
  geom_segment(aes(y = 0, 
                   x = name, 
                   yend = age, 
                   xend= name), 
               color = "black") +
  coord_flip()
```


Loki is throwing things off here. Let's take him out:  
```{r lollipop2}
superAnti %>% filter(name != "Loki") %>% ggplot(aes(x=name, y=age, label=age )) + 
  geom_point(stat='identity', fill="black", size=8) +
  geom_segment(aes(y = 0, 
                   x = name, 
                   yend = age, 
                   xend= name), 
               color = "black") +
  coord_flip()
```
Much better!

Let's add a little text in the dots for clarity and set our 'baseline' age to 32:
```{r lollipop3}
superAnti %>% filter(name != "Loki") %>% ggplot(aes(x=name, y=age, label=age )) + 
  geom_point(stat='identity', fill="black", size=8) +
  geom_segment(aes(y = 32, 
                   x = name, 
                   yend = age, 
                   xend= name), 
               color = "black") +
                geom_text(color="white", size=3)+
  labs(title="Who's older and younger than the Batman?", subtitle="Batman is 32")+
  coord_flip()
```